---
bibliography: Supplemental_file.bib
biblio-style: apalike
highlight_bw: yes
output:
  rmarkdown::pdf_document:
    includes:
        in_header: header.tex 
    dev: "cairo_pdf"
    latex_engine: xelatex
    fig_caption: yes
---

**Supplemental File of**

# ggmsa: a visual exploration tool for multiple sequence alignment and associated data

\renewcommand{\figurename}{Fig.}

\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }

\beginsupplement


```{r message=FALSE, warning=FALSE, echo = FALSE}
library(knitr)
knit_hooks$set(crop = hook_pdfcrop)
opts_chunk$set(crop = TRUE)
```


```{r message=FALSE, warning=FALSE}
library(ggmsa)
library(ggplot2)
library(magick)
library(ggplotify)
library(aplot)
library(RColorBrewer)
library(Biostrings)
library(treeio)
library(ggtree)
```

## Reviewer1: Require2

```{r  fig.height=4, fig.width=8, warning=FALSE, message=FALSE}

tpp_seq <- "data/tpp_riboswitch.fasta"
arc_4NYG <- "data/riboswitch_thiamine.txt"
arc_4NYD <- "data/riboswitch_hypoxanthine.txt"
thiamine <- readSSfile(arc_4NYG, type = "Vienna" )
hypoxanthine <- readSSfile(arc_4NYD, type = "Vienna")

p <- ggmsa(tpp_seq, 
           color = "Chemistry_NT", 
           seq_name = F, 
           show.legend = F, 
           border = NA) +
    geom_helix(helix_data = list(known = hypoxanthine, 
                                 predicted = thiamine)) + 
    theme(axis.text.y = element_blank())

p1 <- image_read_pdf("data/bpRNA_PDB_590_ColorCodedStructures_4NYG.pdf")
p2 <- image_read_pdf("data/bpRNA_PDB_589_ColorCodedStructures_4NYD.pdf")

q1 <- as.ggplot(p1)
q2 <- as.ggplot(p2)

p_loop <- plot_list(gglist = list(q1, q2), ncol = 1)
pp <- plot_list(gglist = list(p_loop, p), ncol = 2, tag_levels = "A")
pp
```

## Reviewer1: Require2

```{r fig.height=5, fig.width=10, warning=FALSE, message=FALSE}
colRD <- colorRampPalette(rev(brewer.pal(n = 9, name = "OrRd")))
colBU <- colorRampPalette(colors = rev(c("#185089","#FFF7EC")))

data <- "data/s_RBD.fasta"
mds <- read.csv("data/MDS.csv")
del <- c("expr_lib1", "expr_lib2", 
         "expr_avg","bind_lib1", 
         "bind_lib2")

mds <- mds[,!colnames(mds) %in% del]

tidymsa <- tidy_msa(data)
tidymsa <- assign_mds(tidymsa, mds)
tidymsa$position <- tidymsa$position + 330 

p <- ggplot() + 
    geom_msa(data = tidymsa,
             char_width = 0.5,
             mds = TRUE, 
             seq_name = TRUE, 
             show.legend = TRUE) + 
    ggmsa:::theme_msa() +
    scale_fill_gradientn(name = "ACE2 binding",
                         colors = c(colRD(75),colBU(25))) +
    facet_msa(50) 
p
```



```{r  fig.height=5, fig.width=10, warning=FALSE, message=FALSE}
#read sequences
x <- readAAStringSet("data/ACE2.fasta")
y <- readAAStringSet("data/Spike_RBD.fasta")

#read protein-protein position
inter1 <- read.csv("data/6m0j_inter.csv")
inter2 <- read.csv("data/7wpb_inter.csv")

t1 <- tidy_msa(x, start = 19)
t2 <- tidy_msa(y, start = 331)
t_merge <- merge_seq(t1, gap = 100, t2)

h1 <- tidy_hdata(100, inter1, t1, t2)
h2 <- tidy_hdata(100, inter2, t1, t2)

ACE2 <- t_merge[t_merge$position %in% c(h1$i,h2$i),]
spike <- t_merge[t_merge$position %in% c(h1$j,h2$j),]
t1_reset <- reset_pos(ACE2)
t2_reset <- reset_pos(spike)
sim_merge <- merge_seq(previous_seq = t1_reset,
                       subsequent_seq = t2_reset,
                       gap = 5, 
                       adjust_name = FALSE)

sim_h1 <- simplify_hdata(h1, sim_merge)
sim_h2 <- simplify_hdata(h2, sim_merge)

#break and label
b <- sim_merge[sim_merge$character != "-", "position"] %>% unique()
l <-  c(inter1$Res.no.1,inter1$Res.no..2, 
        inter2$Res.no.1,inter2$Res.no..2) %>% unique

p_sim <- ggplot() + 
    geom_msa(data = sim_merge,border = NA, char_width = 0.5, seq_name = F) + 
    ggmsa:::theme_msa() + 
    geom_helix(helix_data = list(known = sim_h1, 
                             predicted = sim_h2), 
               overlap = T) + 
     geom_text(mapping = aes(x = b, 
                             y = 0.25, 
                             label = l[order(l)]), 
               size = 3) +
     theme(axis.text.x = element_blank(),
           axis.text.y = element_blank()) + 
     scale_y_discrete(labels = c("6m0j:A(ACE2)", "7wpb:D(ACE2)")) + 
     geom_text(aes(x = c(27.2,27.2), 
                   y = c(1,2), 
                   label = c("6m0j:A(Spike_RBD)", 
                             "7wpb:D(Spike_RBD)")),
               size = 3.5) +
    geom_text(aes(x = c(-1.5,-1.5), 
                  y = c(1,2), 
                  label = c("6m0j:A(ACE2)", 
                            "7wpb:D(ACE2)")),
              size = 3.5)  + xlim(-2,29) 
p_sim
```

## Reviewer2: Require1

```{r  fig.height=6, fig.width=12, warning=FALSE, message=FALSE}
x <- read.phyloxml("data/msa_phyloxml.xml")
p <- ggtree(x, size = 2) 

p1 <- adjust_ally(p, node = 11, sub = FALSE)
tidy1 <- extract_seq(p1)

p2 <- adjust_ally(p, node = 11, sub = TRUE)
tidy2 <- extract_seq(p2)

pp1 <- p1 +  geom_facet(geom = geom_msa, data = tidy1,  panel = 'MSA',
                     font = NULL, color = "Chemistry_NT") + xlim_tree(0.15)

pp2 <- p2 + geom_facet(geom = geom_msa, data = tidy2,  panel = 'MSA',
                       font = NULL, color = "Chemistry_NT") + xlim_tree(0.15)

pp <- plot_list(gglist = list(pp1,pp2), 
                nrow = 2, 
                tag_levels = "A", 
                heights = c(0.4,0.6))
pp
```

## Reviewer2: Require2

```{r  fig.height=11, fig.width=12, warning=FALSE, message=FALSE}
maf <- "data/chr1_KI270707v1_random.txt.maf"
ref = "hg38.chr1_KI270707v1_random"
seq_df <- read_maf(maf)
tidy_df <- tidy_maf_df(seq_df, ref = ref)

ggmaf(data = tidy_df, 
      ref = ref, 
      block_start = 1, 
      block_end = 10, 
      facet_field = 5,
      facet_heights = c(0.4,0.6))
```